---
title: "final_project"
subtitle: "Olivia_Lucia"
format: pdf
editor: visual
editor_options: 
  chunk_output_type: inline
---

```{r}
library(readr)
phyto_data <- read_csv("OL-weeklyphyto-data.csv")
```

```{r}
library(dplyr)
library(ggplot2)
library(tidyverse)

head(phyto_data)
summary(phyto_data)
```

```{r}
phyto_data <- phyto_data %>%
  mutate(
    `Chlorophyll (ug/L)` = as.numeric(`Chlorophyll (ug/L)`),
    `Picoeukaryotes (cells/ml)` = as.numeric(`Picoeukaryotes (cells/ml)`),
    `Nanoeukaryotes (cells/mL)` = as.numeric(`Nanoeukaryotes (cells/mL)`),
    `Cyanobacteria (cells/mL)` = as.numeric(`Cyanobacteria (cells/mL)`)) 
summary(phyto_data)
```

```{r}
phyto_data_clean <- phyto_data %>%
  filter(!is.na(`Chlorophyll (ug/L)`), !is.na(`Picoeukaryotes (cells/ml)`))
```

Is chlorophyll concentration related to the type of micro-organism?

```{r}
phyto_data_clean <- phyto_data_clean %>%
  mutate(Date = as.Date(paste(Year, Month, "01", sep = "-"), format = "%Y-%m-%d"))
```

```{r, fig.cap= "Boxplot comparing chlorophyll concentration across different microorganism types (Cyanobacteria, Nanoeukaryotes, and Picoeukaryotes) measured in cells/mL. Points outside the boxplot represent outliers."}
data_long <- phyto_data_clean %>%
  select(Date, `Chlorophyll (ug/L)`, `Picoeukaryotes (cells/ml)`, `Nanoeukaryotes (cells/mL)`, `Cyanobacteria (cells/mL)`, Location) %>%
  pivot_longer(
    cols = c(`Picoeukaryotes (cells/ml)`, `Nanoeukaryotes (cells/mL)`, `Cyanobacteria (cells/mL)`), 
    names_to = "Micro_organism_type", 
    values_to = "Concentration"
  )
#create the boxplot
ggplot(data_long, aes(x = Micro_organism_type, y = `Chlorophyll (ug/L)`, fill = Micro_organism_type)) +
  geom_boxplot() +
  labs(title = "Chlorophyll Concentration by Microorganism Type",
       x = "Microorganism Type",
       y = "Chlorophyll Concentration", fill = "Microorganism Type") +
  theme_minimal() +
  scale_fill_manual(values = c("Picoeukaryotes (cells/ml)" = "blue", 
                               "Nanoeukaryotes (cells/mL)" = "green", 
                               "Cyanobacteria (cells/mL)" = "orange"))
```

same figure with jitter

```{r, fig.cap="Boxplot illustrating chlorophyll concentration (µg/L) for different microorganism types: Cyanobacteria, Nanoeukaryotes, and Picoeukaryotes, measured in cells/mL. Each point represents an individual measurement, highlighting the potential variability and outliers in chlorophyll concentration among the three microorganism groups."}
# Scatter plot with linear trend lines for each microorganism type
ggplot(data_long, aes(x = Micro_organism_type, y = `Chlorophyll (ug/L)`, color = Micro_organism_type)) +
  geom_jitter() +  # Adds jitter to spread points out horizontally
  geom_boxplot() +  # Adds a boxplot overlay for distribution view
  labs(title = "Chlorophyll Concentration by Micro-organism Type",
       x = "Micro-organism Type", y = "Chlorophyll (ug/L)", color = "Microorganism Type") +
  theme_minimal() +
  scale_color_manual(values = c("Picoeukaryotes (cells/ml)" = "blue", 
                                "Nanoeukaryotes (cells/mL)" = "green", 
                                "Cyanobacteria (cells/mL)" = "orange"))
```

```{r}
data_long <- data_long %>%
  mutate(Date = as.Date(Date, format = "%Y-%m-%d"))
library(lubridate)

data_long <- data_long %>%
  mutate(
    Year = year(Date),
    Month = month(Date)
  )
data_long <- data_long %>% 
  mutate(
    Month = case_when(
      Month %in% c(1) ~ "Jan",
      Month %in% c(2) ~ "Feb",
      Month %in% c(3) ~ "Mar",
      Month %in% c(4) ~ "Apr",
      Month %in% c(5) ~ "May",
      Month %in% c(6) ~ "Jun",
      Month %in% c(7) ~ "Jul",
      Month %in% c(8) ~ "Aug",
      Month %in% c(9) ~ "Sep",
      Month %in% c(10) ~ "Oct",
      Month %in% c(11) ~ "Nov",
      Month %in% c(12) ~ "Dec")) 
```

```{r}
seasonal_data <- data_long %>%
  group_by(Year, Month,Date, Micro_organism_type) %>%
  summarize(Average_Concentration = mean(Concentration)) 

seasonal_data <- seasonal_data %>%
  mutate(
    Date = as.Date(Date),  # Ensure Date is in Date format
    Year = year(Date),
    Month = month(Date),
    Season = case_when(
      Month %in% c(12, 1, 2) ~ "Winter",
      Month %in% c(3, 4, 5) ~ "Spring",
      Month %in% c(6, 7, 8) ~ "Summer",
      Month %in% c(9, 10, 11) ~ "Fall"
    )
  )
```

figure of microorganisms concentration over the 4 year period faceted by season

```{r, fig.cap="Line plots showing the seasonal trends in microorganism density (cells/mL) over time from 2020 to 2024 for Cyanobacteria, Nanoeukaryotes, and Picoeukaryotes. Each panel represents a different season (Fall, Spring, Summer, and Winter)."}
ggplot(seasonal_data, aes(x = Year, y = Average_Concentration, color = Micro_organism_type)) +
  geom_point() +                      
  geom_line(aes(group = Micro_organism_type)) + # Connect points with lines
  facet_wrap(~ Season, scales = "free_y") +   # Facet by Season
  labs(
    title = "Microorganism Density Over Time by Season",
    x = "Year",
    y = "Microorganism Density (cells/mL)",
    color = "Microorganism Type"
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 12),
    panel.grid.minor = element_blank()
  )
```

also look at chloropyll concentration over time by season

```{r}
seasonal_data2 <- data_long %>%
  group_by(Year, Month,Date, `Chlorophyll (ug/L)`, Micro_organism_type, Location) %>%
  summarize(avg_chloropyll = mean(`Chlorophyll (ug/L)`)) 

seasonal_data2 <- seasonal_data2 %>%
  mutate(
    Date = as.Date(Date),  # Ensure Date is in Date format
    Year = year(Date),
    Month = month(Date),
    Season = case_when(
      Month %in% c(12, 1, 2) ~ "Winter",
      Month %in% c(3, 4, 5) ~ "Spring",
      Month %in% c(6, 7, 8) ~ "Summer",
      Month %in% c(9, 10, 11) ~ "Fall"
    )
  )
```

```{r, fig.cap="Boxplots showing the seasonal trends in chloropyll concentration(ug/mL) over time from 2020 to 2024 for. Each panel represents a different season (Fall, Spring, Summer, and Winter)."}
ggplot(seasonal_data2, aes(x = as.factor(Year), y = avg_chloropyll)) + # Convert Year to a factor for categorical plotting
  geom_boxplot() + # Use boxplot instead of points and lines
  facet_wrap(~ Season, scales = "free_y") + # Facet by Season
  labs(
    title = "Chlorophyll Concentrations Over Time by Season",
    x = "Year",
    y = "Concentration (µg/mL)"
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 12),
    panel.grid.minor = element_blank()
  )
```

figure faceted by month looking at microorganisms concentrations over the 4 year period

faceted by mircoorganisms type and looking at the density of

```{r, fig.cap="Line plots showing the trends in microorganism density (cells/mL) over time from 2020 to 2024 for Cyanobacteria, Nanoeukaryotes, and Picoeukaryotes. Each panel represents a different microorganism."}
ggplot(seasonal_data, aes(x = Year, y = Average_Concentration, color = Month)) +
  geom_line() +
  facet_wrap(~Micro_organism_type, scales = "free_y") +
  labs(title = "Variations in Phytoplankton Density by Microorganism Type Over time",
       x = "Year",
       y = "Average Phytoplankton Density") +
  theme_minimal()
```

do this for chlorophyll instead of density too

```{r, fig.cap="Line plots showing the trends in chlorophyll concentration (ug/mL) over time from 2020 to 2024 for Cyanobacteria, Nanoeukaryotes, and Picoeukaryotes. Each panel represents a different microorganism."}
ggplot(seasonal_data2, aes(x = Year, y = avg_chloropyll, color = Month)) +
  geom_line() +
  facet_wrap(~Micro_organism_type, scales = "free_y") +
  labs(title = "Variations in Chlorophyll Concentration by Microorganism Type Over time",
       x = "Year",
       y = "Average Chlorophyll concentration") +
  theme_minimal()
```

look at chlorphyll concnetration over the 4 year period by location

Linear regression between chlorophyll and picoeukaryotes

```{r, fig.cap="Relationship between chlorophyll concentration and picoeukaryotes density, including best fit line."}
lm_model <- lm(`Chlorophyll (ug/L)` ~ `Picoeukaryotes (cells/ml)`, data = phyto_data_clean)
summary(lm_model)

# Plot with regression line
ggplot(phyto_data_clean, aes(x = `Picoeukaryotes (cells/ml)`, y = `Chlorophyll (ug/L)`)) +
  geom_point() +
  geom_smooth(method = "lm", col = "blue") +
  labs(title = "Chlorophyll vs. Picoeukaryotes", x = "Picoeukaryotes (cells/ml)", y = "Chlorophyll (ug/L)")

```


