---
title: "Multi-Year Analysis of Phytoplankton Community Structure Across UNH Marine Laboratories"
author: "Olivia Lucia"
date: "December 6, 2024"
format: pdf
pdf-engine: xelatex  # Or "lualatex"
mainfont: "Times New Roman"
---

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(tidyverse)
```
## GitHub Repository Link:

https://github.com/obl1001/BIOL806Final.git

## Introduction

Phytoplankton, a varied collection of photosynthetic microorganisms, are the building blocks of aquatic ecosystems and play an important role in preserving global ecological equilibrium (Pomeroy, Paffenhöfer, & Yoder, 1987). These bacteria promote primary production in seas, lakes, and rivers, forming the foundation of the aquatic food web. They account for about half of Earth's photosynthesis, making them important participants in carbon sequestration and nutrient cycle (Polimene et al., 2017). Phytoplankton facilitates the movement of energy from lower trophic levels, such as zooplankton, to higher trophic levels, including fish and other marine species. Furthermore, their contributions to biogeochemical processes such as the nitrogen and phosphorus cycles demonstrate their importance in maintaining ecosystem health and production (Pomeroy, Paffenhöfer, & Yoder, 1987).

Chlorophyll content, which is commonly employed as a proxy for phytoplankton biomass and production, gives useful information about their abundance and ecological health. However, the variability in chlorophyll levels among phytoplankton groups, as well as the environmental conditions that drive these variations, are poorly understood (Riemann, Simonsen, & Stensgaard, 1989). Light availability, temperature, and nutrient concentrations all have an impact on phytoplankton abundance, as well as the makeup of distinct groups within the community. These groupings, which include picoeukaryotes, nanoeukaryotes, and cyanobacteria, inhabit discrete ecological niches and have diverse physiological characteristics, making them sensitive indicators of environmental change.

Environmental variables influence phytoplankton community composition in complicated ways. Light availability, for example, determines photosynthetic activity and vertical distribution in the water column. Nutrient availability, particularly nitrogen and phosphorus, influences growth rates and biomass output (Gattuso et al., 2006). Temperature, on the other hand, affects enzyme activity and cellular metabolism, influencing the prevalence of certain phytoplankton types. Seasonal patterns, such as spring blooms in temperate locations, show how fluctuations in light and nutrient availability cause dramatic increases in phytoplankton abundance (May et al., 2003). Climate events like El Niño-Southern Oscillation (ENSO) can produce interannual fluctuations in temperature, nutrient input, and mixing regimes, complicating these processes even more.

Among the several types of phytoplankton, picoeukaryotes and nanoeukaryotes play critical roles in microbial food chains. Picoeukaryotes, the smallest eukaryotic phytoplankton, thrive in nutrient-poor settings because of their large surface area-to-volume ratio, which allows for efficient nutrient absorption (Blachinsky, 2024). Nanoeukaryotes, which are somewhat bigger than picoeukaryotes, provide considerable contributions to primary production and act as an intermediary between tiny phytoplankton and larger grazers. Cyanobacteria, sometimes known as blue-green algae, are unusual among phytoplankton because they can fix atmospheric nitrogen, allowing them to grow in nutrient-depleted seas (Gupta, Prakash, & Gupta, 2015).

Temporal patterns in phytoplankton abundance provide insights into ecosystem dynamics and reactions to environmental change. Short-term variations, like as daily changes in light intensity, have an impact on photosynthetic activity and nutrient intake. Seasonal variations in temperature, light, and nutrition availability cause predictable patterns in phytoplankton abundance, such as spring and fall blooms (Côté & Platt, 1983). Long-term trends are more complicated, reflecting the cumulative effects of climate change, nutrient enrichment, and other human-caused causes. Warming sea surface temperatures and changed stratification patterns, for example, might disrupt typical seasonal cycles, favoring certain phytoplankton species over others and potentially lowering total productivity (Klauschies et al., 2012).

Chlorophyll concentration, as an integrative indicator of phytoplankton biomass, is an essential tool for tracking these developments. Researchers can get insights into the mechanisms governing community composition and production by studying the connections between chlorophyll levels and the densities of distinct phytoplankton groups (Riemann, Simonsen, & Stensgaard, 1989). Understanding phytoplankton dynamics is critical not just for ecological study, but also for tackling larger environmental issues. As climate change increases, changes in phytoplankton populations might have far-reaching consequences for marine food webs, fisheries, and carbon cycling (CHANGE, 2019). Changes in primary production and nutrient cycling may impact the ocean's ability to trap carbon, worsening global warming. Furthermore, imbalances in phytoplankton populations, such as toxic algal blooms, endanger human health, marine biodiversity, and coastal economy (Garmendia et al., 2011). This research can help to guide management methods for protecting ecosystem services and minimizing the effects of environmental change by identifying the variables that influence phytoplankton abundance and community composition.

For this study a four-year dataset was utilized to investigate the links between chlorophyll concentration, phytoplankton groups, and temporal variability. By concentrating on picoeukaryotes, nanoeukaryotes, and cyanobacteria, we hope to discover patterns in their abundance and how they relate to changes in environmental circumstances. Furthermore, we investigate the significance of these patterns for ecosystem stability and production, especially in light of ongoing climate change. This study not only increases our understanding of phytoplankton ecology, but also emphasizes the significance of monitoring and protecting these minute yet critical components of aquatic ecosystems. Focusuing on two main research questions: 1) Is chlorophyll concentration linked to certain types of microorganisms? 2) How do the concentrations of various phytoplankton groups change over a four-year period? Addressing these concerns will help us get a better knowledge of phytoplankton ecology, namely the interactions between community composition, environmental factors, and temporal dynamics.


```{r pulling data, include=FALSE}
library(readr)
phyto_data <- read_csv("OL-weeklyphyto-data.csv")
```


```{r data cleaning, include=FALSE}
#Mutating columns 
phyto_data <- phyto_data %>%
  mutate(
    `Chlorophyll (ug/L)` = as.numeric(`Chlorophyll (ug/L)`),
    `Picoeukaryotes (cells/ml)` = as.numeric(`Picoeukaryotes (cells/ml)`),
    `Nanoeukaryotes (cells/mL)` = as.numeric(`Nanoeukaryotes (cells/mL)`),
    `Cyanobacteria (cells/mL)` = as.numeric(`Cyanobacteria (cells/mL)`)) 
summary(phyto_data)

#Filtering
phyto_data_clean <- phyto_data %>%
  filter(!is.na(`Chlorophyll (ug/L)`), !is.na(`Picoeukaryotes (cells/ml)`))

#Is chlorophyll concentration related to the type of micro-organism?

#Adding Date Column
phyto_data_clean <- phyto_data_clean %>%
  mutate(Date = as.Date(paste(Year, Month, "01", sep = "-"), format = "%Y-%m-%d"))
```

## Methods

The data set analyzed in this study comprises weekly samples of chlorophyll concentration and densities of various phytoplankton types, collected from aquatic environments at Hampton Beach, Coastal Marine Laboratory at Jaffrey Point, and Jackson Estuary Marine Laboratory at Adam’s Point. Sampling occurred at regular intervals to capture seasonal and temporal variations in phytoplankton communities. Key variables include chlorophyll concentration (µg/L), representing phytoplankton biomass, and densities of microorganism types such as picoeukaryotes, nanoeukaryotes, and cyanobacteria. Temporal variables (month and year) were included to enable the analysis of seasonal and long-term trends.

All data wrangling and analyses were conducted in R v. 4.4.1 (R Core Team, 2020). Initial data cleaning addressed inconsistencies, missing values, and outliers to ensure data quality. The data set was reshaped using the pivot_longer() function, converting wide-format phytoplankton density columns into a long format for streamlined comparisons across microorganism types.

To investigate the relationship between chlorophyll concentration and microorganism types, exploratory data analysis included visualizations such as box plots (chlorophyll concentration on the y-axis and microorganism type on the x-axis) and scatter plots with jitter to capture variability within groups. These visualizations revealed patterns and associations between chlorophyll levels and microorganism categories.

Temporal variations in phytoplankton density were analyzed by aggregating data by month and year. The average density of each microorganism group was calculated for each time period. Seasonal trends were examined using line plots for each phytoplankton group, with the x-axis representing years and the y-axis depicting microbe density. These plots provided insights into recurrent seasonal peaks or troughs and long-term density fluctuations.

## Results and Analysis

```{r, echo=FALSE, fig.cap= "Boxplot comparing chlorophyll concentration across different microorganism types (Cyanobacteria, Nanoeukaryotes, and Picoeukaryotes) measured in cells/mL. Points outside the boxplot represent outliers."}
data_long <- phyto_data_clean %>%
  select(Date, `Chlorophyll (ug/L)`, `Picoeukaryotes (cells/ml)`, `Nanoeukaryotes (cells/mL)`, `Cyanobacteria (cells/mL)`, Location) %>%
  pivot_longer(
    cols = c(`Picoeukaryotes (cells/ml)`, `Nanoeukaryotes (cells/mL)`, `Cyanobacteria (cells/mL)`), 
    names_to = "Micro_organism_type", 
    values_to = "Concentration"
  )
#create the boxplot
ggplot(data_long, aes(x = Micro_organism_type, y = `Chlorophyll (ug/L)`, fill = Micro_organism_type)) +
  geom_boxplot() +
  labs(
    title = "Chlorophyll Concentration by Microorganism Type",
    x = "Microorganism Type",
    y = "Chlorophyll (ug/L)", 
    fill = "Microorganism Type"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1) # Rotate x-axis text for readability
  ) +
  scale_fill_manual(
    values = c(
      "Picoeukaryotes (cells/ml)" = "blue", 
      "Nanoeukaryotes (cells/mL)" = "green", 
      "Cyanobacteria (cells/mL)" = "orange"
    )
  )
```

```{r jitter, echo=FALSE, fig.cap="Boxplot illustrating chlorophyll concentration (µg/L) for different microorganism types: Cyanobacteria, Nanoeukaryotes, and Picoeukaryotes, measured in cells/mL. Each point represents an individual measurement, highlighting the potential variability and outliers in chlorophyll concentration among the three microorganism groups."}
# Scatter plot with linear trend lines for each microorganism type
ggplot(data_long, aes(x = Micro_organism_type, y = `Chlorophyll (ug/L)`, color = Micro_organism_type)) +
  geom_jitter() +  # Adds jitter to spread points out horizontally
  geom_boxplot() +  # Adds a boxplot overlay for distribution view
  labs(title = "Chlorophyll Concentration by Micro-organism Type",
       x = "Micro-organism Type", y = "Chlorophyll (ug/L)", color = "Microorganism Type") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1) # Rotate x-axis text for readability
  ) +
  scale_color_manual(values = c("Picoeukaryotes (cells/ml)" = "blue", 
                                "Nanoeukaryotes (cells/mL)" = "green", 
                                "Cyanobacteria (cells/mL)" = "orange"))
```

*Chlorophyll Concentration Amoung Microorganisms* 

Figures 1 and 2 show boxplots demonstrating the lack of heterogeneity in chlorophyll content among the three microorganism kinds. All three species of bacteria have similar, nearly equal chlorophyll levels. Picoeukaryotes had a significantly wider range of concentrations, but with occasional outliers, as shown by the jittered data points in Figure 2.

```{r density, include=FALSE}
data_long <- data_long %>%
  mutate(Date = as.Date(Date, format = "%Y-%m-%d"))
library(lubridate)

data_long <- data_long %>%
  mutate(
    Year = year(Date),
    Month = month(Date)
  )
data_long <- data_long %>% 
  mutate(
    Month = case_when(
      Month %in% c(1) ~ "Jan",
      Month %in% c(2) ~ "Feb",
      Month %in% c(3) ~ "Mar",
      Month %in% c(4) ~ "Apr",
      Month %in% c(5) ~ "May",
      Month %in% c(6) ~ "Jun",
      Month %in% c(7) ~ "Jul",
      Month %in% c(8) ~ "Aug",
      Month %in% c(9) ~ "Sep",
      Month %in% c(10) ~ "Oct",
      Month %in% c(11) ~ "Nov",
      Month %in% c(12) ~ "Dec")) 

seasonal_data <- data_long %>%
  group_by(Year, Month,Date, Micro_organism_type) %>%
  summarize(Average_Concentration = mean(Concentration)) 

seasonal_data <- seasonal_data %>%
  mutate(
    Date = as.Date(Date),  # Ensure Date is in Date format
    Year = year(Date),
    Month = month(Date),
    Season = case_when(
      Month %in% c(12, 1, 2) ~ "Winter",
      Month %in% c(3, 4, 5) ~ "Spring",
      Month %in% c(6, 7, 8) ~ "Summer",
      Month %in% c(9, 10, 11) ~ "Fall"
    )
  )
```


```{r line plot, echo=FALSE, fig.cap="Line plots showing the seasonal trends in microorganism density (cells/mL) over time from 2020 to 2024 for Cyanobacteria, Nanoeukaryotes, and Picoeukaryotes. Each panel represents a different season (Fall, Spring, Summer, and Winter)."}

ggplot(seasonal_data, aes(x = Year, y = Average_Concentration, color = Micro_organism_type)) +
  geom_point() +                      
  geom_line(aes(group = Micro_organism_type)) + # Connect points with lines
  facet_wrap(~ Season, scales = "free_y") +   # Facet by Season
  labs(
    title = "Microorganism Density Over Time by Season",
    x = "Year",
    y = "Microorganism Density (cells/mL)",
    color = "Microorganism Type"
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1), # Rotate x-axis labels
    panel.grid.minor = element_blank()
  )
```


```{r seasons, include=FALSE}
#Seasonal data number 2
seasonal_data2 <- data_long %>%
  group_by(Year, Month,Date, `Chlorophyll (ug/L)`, Micro_organism_type, Location) %>%
  summarize(avg_chloropyll = mean(`Chlorophyll (ug/L)`)) 

seasonal_data2 <- seasonal_data2 %>%
  mutate(
    Date = as.Date(Date),  # Ensure Date is in Date format
    Year = year(Date),
    Month = month(Date),
    Season = case_when(
      Month %in% c(12, 1, 2) ~ "Winter",
      Month %in% c(3, 4, 5) ~ "Spring",
      Month %in% c(6, 7, 8) ~ "Summer",
      Month %in% c(9, 10, 11) ~ "Fall"
    )
  )
```

```{r boxplot, echo=FALSE, fig.cap="Boxplots showing the seasonal trends in chloropyll concentration(ug/mL) over time from 2020 to 2024 for. Each panel represents a different season (Fall, Spring, Summer, and Winter)."}
ggplot(seasonal_data2, aes(x = as.factor(Year), y = avg_chloropyll)) + # Convert Year to a factor for categorical plotting
  geom_boxplot() + # Use boxplot instead of points and lines
  facet_wrap(~ Season, scales = "free_y") + # Facet by Season
  labs(
    title = "Chlorophyll Concentrations Over Time by Season",
    x = "Year",
    y = "Concentration (µg/mL)"
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 12),
    panel.grid.minor = element_blank()
  )
```

*Seasonal Trends in Microorganism density and chlorophyll concentration* 

Seasonal line plot (Figures 3) and boxplot (Figure 4) showed unique patterns across time. Microorganism density varied seasonally, with Cyanobacteria and Picoeukaryotes peaking in the spring and summer, whereas Nanoeukaryotes remained more steady throughout the year. Chlorophyll concentrations paralleled similar patterns, indicating a possible relationship between microbe abundance and chlorophyll levels.

```{r,echo=FALSE, fig.cap="Line plots showing the trends in microorganism density (cells/mL) over time from 2020 to 2024 for Cyanobacteria, Nanoeukaryotes, and Picoeukaryotes. Each panel represents a different microorganism."}
ggplot(seasonal_data, aes(x = Year, y = Average_Concentration, color = Month)) +
  geom_line() +
  facet_wrap(
    ~ Micro_organism_type,
    scales = "free_y",
    labeller = label_wrap_gen(width = 15) # Adjust the width for better wrapping
  ) +
  labs(
    title = "Variations in Phytoplankton Density by Microorganism Type Over Time",
    x = "Year",
    y = "Average Phytoplankton Density"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1), # Rotate x-axis labels
   )
```

```{r,echo=FALSE, fig.cap="Line plots showing the trends in chlorophyll concentration (ug/mL) over time from 2020 to 2024 for Cyanobacteria, Nanoeukaryotes, and Picoeukaryotes. Each panel represents a different microorganism."}
ggplot(seasonal_data2, aes(x = Year, y = avg_chloropyll, color = Month)) +
  geom_line() +
  facet_wrap(
    ~ Micro_organism_type,
    scales = "free_y",
    labeller = label_wrap_gen(width = 15) # Wrap long facet titles
  ) +
  labs(
    title = "Variations in Chlorophyll Concentration by Microorganism Type Over Time",
    x = "Year",
    y = "Average Chlorophyll Concentration"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1), # Rotate x-axis labels for readability
   )
```

*Long-term trends by microorganism Type* 

Figures 5 and 6 show trends in density and chlorophyll concentration for each microbe throughout time. Figure 5 demonstrated a lot of fluctuation in microbe density, with no pattern or association. Chlorophyll concentrations followed similar trajectories, with all three microorganisms having nearly identical chlorophyll concentration tendencies, as seen in Figures 1 and 2.

```{r lm, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Relationship between chlorophyll concentration and picoeukaryotes density, including best fit line."} 
# Create the linear model
lm_model <- lm(`Chlorophyll (ug/L)` ~ `Picoeukaryotes (cells/ml)`, data = phyto_data_clean)

# Plot with regression line
ggplot(phyto_data_clean, aes(x = `Picoeukaryotes (cells/ml)`, y = `Chlorophyll (ug/L)`)) +
  geom_point() +
  geom_smooth(method = "lm", col = "blue") +
  labs(title = "Chlorophyll vs. Picoeukaryotes", x = "Picoeukaryotes (cells/ml)", y = "Chlorophyll (ug/L)")
```

*Correlation between chlorophyll and picoeukaryote density* 

A linear regression analysis (Figure 7) found a slightly negative connection between chlorophyll concentration and Picoeukaryote density. This demonstrates that, despite their low overall density, picoeukaryotes make large contributions to chlorophyll production, possibly due to their efficient photosynthetic activities. These findings emphasize phytoplankton's diverse contributions to ecosystem output, as well as the importance of seasonal and long-term dynamics in phytoplankton population formation.

## Disscussion
From 2020 to 2024, this study looked at the relationships between chlorophyll content, microorganism density, and temporal changes in three major phytoplankton groups: cyanobacteria, nanoeukaryotes, and picoeukaryotes. The findings provide light on the complicated dynamics of phytoplankton communities, emphasizing their heterogeneity across time periods and ecological relevance.

*Chlorophyll Concentration Among Microorganisms*

Interestingly, no significant changes in chlorophyll concentrations were found across the three phytoplankton groups, despite their diverse ecological responsibilities and physiological properties (Figures 1 and 2). When averaged across environmental circumstances, this observation shows that chlorophyll content is very consistent. However, picoeukaryotes have a far larger range of chlorophyll contents, with a few outliers. This diversity is consistent with their ecological adaptability; picoeukaryotes may live in a variety of habitats, from nutrient-poor oligotrophic seas to nutrient-rich coastal zones. Their excellent photosynthetic apparatus and plasticity may allow them to maximize chlorophyll production under changing circumstances, explaining the wider range seen.

This observation raises issues regarding how various phytoplankton communities contribute to primary production. For example, whereas picoeukaryotes are less numerous overall, their high photosynthetic efficiency implies that they may play a disproportionate role in chlorophyll production, especially in nutrient-limited situations (Blachinsky, 2024). This highlights the need for more research into the physiological and genetic factors influencing chlorophyll variations within and among phytoplankton species.

*Seasonal Trends in Microorganism Density and Chlorophyll Concentration*

Seasonal patterns emerged in the data, demonstrating how environmental cycles influence phytoplankton populations. Cyanobacteria and Picoeukaryotes showed density maxima in the spring and summer, coinciding with increasing light availability and nutrient upwelling (Figures 3 and 4). These findings are consistent with prior research, which has shown that phytoplankton blooms frequently occur during the seasons when photosynthesis and growth conditions are greatest (May et al., 2003). Cyanobacteria, in particular, may benefit from their capacity to fix nitrogen, allowing them to survive even in situations where other nutrients, such as nitrate, have been reduced during summer stratification (Gupta, Prakash, & Gupta, 2015).

Nanoeukaryotes, on the other hand, had a more steady density across the year. This stability may represent their greater ecological tolerance or capacity to use a larger range of nutritional sources, making them less subject to seasonal variations. Their consistent presence demonstrates their importance as a stabilizing component of the phytoplankton community, maintaining ecosystem function under a variety of environmental situations (Garmendia et al., 2011).

Chlorophyll concentrations tracked seasonal abundance fluctuations, supporting the close link between microbial density and chlorophyll content. The synchronization of these tendencies emphasizes the importance of seasonal cycles in controlling phytoplankton production and biomass (Riemann, Simonsen, & Stensgaard, 1989). However, it raises concerns regarding possible latencies or mismatches between chlorophyll synthesis and microbial density, which might be investigated in future research.

*Long-term Trends in Microorganism Density and Chlorophyll Concentration*

Analysis of long-term trends found considerable changes in microbe density over a four-year period, but no apparent directional patterns appeared. These findings indicate that the interaction of environmental variables like as temperature, nutrient availability, and interspecies competition results in a dynamic equilibrium in phytoplankton communities. The observed fluctuation may be due to the effect of episodic events like storms, heatwaves, or nutrient pulses, which can either hinder or increase phytoplankton development (Klauschies et al., 2012).

Similarly, chlorophyll concentrations fluctuated in parallel across all three groups, indicating some degree of interconnection among the phytoplankton population. This interconnection underlines the possibility of collective reactions to environmental factors, as well as competitive or facilitative interactions among groupings. For example, changes in food availability may favor one group over another, thereby impacting chlorophyll levels throughout the population.

The lack of a distinct long-term directional trend in chlorophyll concentrations might indicate that growth and loss mechanisms—such as predation, viral lysis, and sinking—are in equilibrium throughout this time period (Pomeroy, Paffenhöfer, & Yoder, 1987). However, it is possible that exogenous causes, such as climate change, have not yet caused noticeable modifications in these ecosystems, or that the study's period is insufficient to capture such changes. Long-term monitoring will be required to detect developing patterns and understand how phytoplankton populations adapt to changing environmental conditions.

*Correlation Between Chlorophyll and Picoeukaryote Density*

The somewhat negative connection found between chlorophyll content and picoeukaryote density (Figure 7) is quite fascinating. Despite having a lower total density than other groupings, picoeukaryotes appear to contribute disproportionately to chlorophyll production. This shows that they may have very efficient photosynthetic systems, allowing them to remain productive even in low-density environments (Hernández-Ruiz et al., 2020).

This observation emphasizes the variety of phytoplankton communities and the need of investigating group-specific contributions to primary production. It also raises doubts regarding picoeukaryotes' ecological tactics, namely whether their high chlorophyll concentration is an adaptive response to light-limited conditions or a result of other physiological processes (Hernández-Ruiz et al., 2020). Future studies might look at these mechanisms further, perhaps revealing new information on the significance of picoeukaryotes in marine ecosystems.

*Implications and Future Directions*

The findings of this study provide important insights into the dynamics of phytoplankton communities, offering light on the links between density, chlorophyll concentration, and temporal variability. Seasonal patterns underscore the role of regular environmental cycles in influencing community composition and production, whereas long-term variability shows the impact of complex and episodic causes. These trends highlight the importance of continued monitoring to better understand the causes of phytoplankton dynamics and how they respond to environmental change.

Future study should expand on these findings by investigating the physiological and genetic factors that govern chlorophyll variations among phytoplankton groups. For example, transcriptome or proteomic investigations might show how picoeukaryotes and other species regulate photosynthesis in different environments. Furthermore, research into interactions between phytoplankton groups, such as competition for resources or facilitation through nutrient recycling, might give more insight into community dynamics.

The ecological consequences of these studies are vast. As climate change modifies temperature regimes, stratification patterns, and nutrient availability, phytoplankton populations may fluctuate in composition and productivity, affecting marine food webs and biogeochemical cycles. Understanding the baseline dynamics of these communities is critical for anticipating and reducing the effects of such changes. Furthermore, the function of picoeukaryotes and other understudied taxa in sustaining primary production poorly varying conditions emphasizes the need of phytoplankton variety as a buffer against environmental change.

Finally, this work improves our understanding of phytoplankton ecology by revealing the links between chlorophyll content, microorganism density, and temporal variability. By merging seasonal, long-term, and group-specific viewpoints, it lays the groundwork for future study into phytoplankton's complex and dynamic functions in marine ecosystems. The findings presented highlight the value of phytoplankton monitoring as a tool for measuring ecosystem health and resilience in the face of continuing environmental change.

## References 

Blachinsky, A. (2024). Inorganic & Organic Nitrogen Uptake by Phytoplankton During Periods of Strongly Contrasting Nitrogen Availability in the Eastern Mediterranean Sea (Master's thesis, University of Haifa (Israel)).  

CHANGE, G. (2019). Scientists’ warning to humanity: microorganisms and climate change.  

Côté, B., & Platt, T. (1983). Day‐to‐day variations in the spring‐summer photosynthetic parameters of coastal marine phytoplankton. Limnology and Oceanography, 28(2), 320-344.  

Garmendia, M., Revilla, M., Bald, J., Franco, J., Laza-Martínez, A., Orive, E., ... & Borja, Á. (2011). Phytoplankton communities and biomass size structure (fractionated chlorophyll “a”), along trophic gradients of the Basque coast (northern Spain). Biogeochemistry, 106, 243-263.  

Gattuso, J. P., Gentili, B., Duarte, C. M., Kleypas, J. A., Middelburg, J. J., & Antoine, D. (2006). Light availability in the coastal ocean: impact on the distribution of benthic photosynthetic organisms and their contribution to primary production. Biogeosciences, 3(4), 489-513.  

Gupta, C., Prakash, D., & Gupta, S. (2015). Role of blue green algae in environment management. Environmental Microbiology. IK International Publishing House Pvt. Ltd., India.  

Hernández-Ruiz, M., Barber-Lluch, E., Prieto, A., Logares, R., & Teira, E. (2020). Response of pico-nano-eukaryotes to inorganic and organic nutrient additions. Estuarine, Coastal and Shelf Science, 235, 106565.  

Klauschies, T., Bauer, B., Aberle-Malzahn, N., Sommer, U., & Gaedke, U. (2012). Climate change effects on phytoplankton depend on cell size and food web structure. Marine Biology, 159, 2455-2478.  

May, C. L., Koseff, J. R., Lucas, L. V., Cloern, J. E., & Schoellhamer, D. H. (2003). Effects of spatial and temporal variability of turbidity on phytoplankton blooms. Marine Ecology Progress Series, 254, 111-128.  

Polimene, L., Sailley, S., Clark, D., Mitra, A., & Allen, J. I. (2017). Biological or microbial carbon pump? The role of phytoplankton stoichiometry in ocean carbon sequestration. Journal of Plankton Research, 39 (2), 180-186.  

Pomeroy, L. R., Paffenhöfer, G. A., & Yoder, J. A. (1987). Interactions of phytoplankton, zooplankton and microorganisms. Progress in Oceanography, 19 (3-4), 353-372.  

Riemann, B., Simonsen, P., & Stensgaard, L. (1989). The carbon and chlorophyll content of phytoplankton from various nutrient regimes. Journal of Plankton Research, 11 (5), 1037-1045. 
